---
title: "Dynamic Assortative Mating - Exploratory Data Analysis"
author: "Linh Nguyen"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    code_folding: 'hide'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# I. Meta-Data

In order to replicate this code file, the following files are needed in the directory `data/` found within the project directory: `dict.xlsx`, `2024-01-15/baseline_cleaned_2024-01-15.RDS`, `2024-01-15/esm_cleaned_2024-01-15.RDS`. R code associated with each code chunk can be viewed and hidden by toggling on `Code` on the top right of each output.

```{r library, results = 'hide', message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(digits = 3)

libraries <- c("apaTables",    # for cor table
               "here",         # directory access
               "psych",        # for scale alphas
               "stringr",      # working with strings
               "tidyr",        # pivot_wider
               "forcats",      # fct_reorder for barplot
               "rio",          # import export
               "summarytools", # descriptives
               "ggplot2",      # plots
               "ggtext",       # element_markdown
               "kableExtra",   # scroll box
               "sjPlot",       # plot_grid
               "dplyr")        # general wrangling

lapply(libraries, require, character.only = TRUE)

bl_path   <- paste0(here(), "/data/2024-01-15/baseline_cleaned_2024-01-15.RDS")
esm_path  <- paste0(here(), "/data/2024-01-15/esm_cleaned_2024-01-15.RDS")
dict_path <- paste0(here(), "/data/dict.xlsx")

source(paste0(here(), "/script/00_helpers.R"))
```

```{r session-info, message = FALSE, warning = FALSE, results = "hold"}
# list of loaded packages and versions
si <- devtools::session_info()[[2]]
rownames(si) <- NULL
si <- as.data.frame(si)
si %>% 
  select(package, loadedversion, date, source) %>% 
  knitr::kable(caption = "All loaded packages") %>%
  kable_styling() %>% 
  scroll_box(height = "300px")
```

```{r data}
# load in data files
baseline <- rio::import(bl_path)
esm      <- rio::import(esm_path)
dict     <- rio::import(dict_path)

# create p_num
baseline$P_num <- as.numeric(
  substr(baseline$Participant_ID, 5, 8)
)
esm$P_num <- as.numeric(
  substr(esm$Participant_ID, 5, 8)
)

# look at top few rows
head(baseline) %>%
  knitr::kable(caption = "Example baseline data") %>%
  kable_styling() %>%
  scroll_box(width = "900px")
head(esm %>% filter(Participant_ID == "1002001") %>% arrange(time_idx)) %>%
  knitr::kable(caption = "Example ESM data") %>%
  kable_styling() %>%
  column_spec(4, width_min = "1in") %>%
  scroll_box(width = "900px")
```

# II. Baseline Data

## 1. Scale Items and Labels

```{r scales, results='asis'}
# self-reported bfas domains and panas
for(iscale in c("Agreeableness", "Conscientiousness",
                "Extraversion", "Neuroticism", "Openness Domain",
                "Positive Affect", "Negative Affect")) {
  
  # grab scale items in the dictionary
  self_dict <- dict %>% 
    filter(data == "baseline" &
           scale == paste("Self", iscale)) %>%
    select(var, var_label, value_label, subscale, key)
  partner_dict <- dict %>% 
    filter(data == "baseline" &
           scale == paste("Partner", iscale))
  
  # calculate alpha
  self_alpha <- round(
    psych::alpha(baseline[self_dict$var])$total["raw_alpha"], 
    2)
  partner_alpha <- round(
    psych::alpha(baseline[partner_dict$var])$total["raw_alpha"], 
    2)
  
  # print dictionary
  self_dict %>%
    knitr::kable(caption = paste("Scale items for", iscale,
                                 "- Cronbach's alpha self =", self_alpha,
                                 "and partner =", partner_alpha)) %>%
    kable_styling() %>%
    scroll_box(height = "300px") %>%
    print()
}

# csi
# grab scale items in the dictionary
current_dict <- dict %>% 
  filter(data == "baseline" &
         scale == "Relationship Quality") %>%
  select(var, var_label, value_label, key)

# calculate alpha
current_alpha <- round(
  psych::alpha(baseline[self_dict$var])$total["raw_alpha"], 
  2)

# print dictionary
current_dict %>%
  knitr::kable(caption = paste("Scale items for Relationship Quality",
                               "- Cronbach's alpha =", current_alpha)) %>%
  kable_styling() %>%
  scroll_box(height = "300px")
```

## 2. Demographics

```{r demo-descr}
# continuous demographics
baseline %>%
  select(age, sex, duration) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of demographics for all participants") %>%
  kable_styling()

baseline %>%
  filter(P_num == 1) %>%
  select(age, sex, duration) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of demographics for first participants") %>%
  kable_styling()
baseline %>%
  filter(P_num == 2) %>%
  select(age, sex, duration) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of demographics for second participants") %>%
  kable_styling()

# categorical demographics
dict %>%
  filter(data == "baseline" &
         var %in% c("sex", "gender", "race_cat", 
                    "edu", "ses", "politics", "religion")) %>%
  select(var, var_label, value_label) %>%
  knitr::kable(caption = "Coding for categorical demographic variables") %>%
  kable_styling()

for(inum in c(1,2)){
  print(paste("Participant", inum))
  baseline %>%
    filter(P_num == inum) %>%
    select(gender, race_cat, edu, ses, politics, religion) %>%
    freq() %>%
    print()
}
```

```{r demo-plots}
p <- plot_hist(var = "age", 
               var_name = "Age", 
               data = baseline, bin_width = 1)

if(signif(var = "age",
          data = baseline)) {
  p <- p + labs(subtitle = "Significant participant difference (p < 0.05)")
  }
print(p)

p <- plot_hist(var = "duration", 
               var_name = "Relationship Duration (in months)", 
               data = baseline, bin_width = 10)

if(signif(var = "duration",
          data = baseline)) {
  p <- p + labs(subtitle = "Significant participant difference (p < 0.05)")
  }
print(p)
```

## 3. Scale Scores

```{r scale-descr}
# BFAS self
baseline %>%
  select(all_of(paste0("self_", 
                       c("agree", "consci", "extra", "neuro", "open")))) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of self-reported BFAS domains for all participants") %>%
  kable_styling()

baseline %>%
  filter(P_num == 1) %>%
  select(all_of(paste0("self_", 
                       c("agree", "consci", "extra", "neuro", "open")))) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of self-reported BFAS domains for first participants") %>%
  kable_styling()
baseline %>%
  filter(P_num == 2) %>%
  select(all_of(paste0("self_", 
                       c("agree", "consci", "extra", "neuro", "open")))) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of self-reported BFAS domains for second participants") %>%
  kable_styling()

# BFAS other
baseline %>%
  select(all_of(paste0("partner_", 
                       c("agree", "consci", "extra", "neuro", "open")))) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of partner-reported BFAS domains for all participants") %>%
  kable_styling()

baseline %>%
  filter(P_num == 1) %>%
  select(all_of(paste0("partner_", 
                       c("agree", "consci", "extra", "neuro", "open")))) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of partner-reported BFAS domains for first participants") %>%
  kable_styling()
baseline %>%
  filter(P_num == 2) %>%
  select(all_of(paste0("partner_", 
                       c("agree", "consci", "extra", "neuro", "open")))) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of partner-reported BFAS domains for second participants") %>%
  kable_styling()

# PANAS
baseline %>%
  select(self_pa, partner_pa, self_na, partner_na) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of PANAS for all participants") %>%
  kable_styling()

baseline %>%
  filter(P_num == 1) %>%
  select(self_pa, partner_pa, self_na, partner_na) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of PANAS for first participants") %>%
  kable_styling()
baseline %>%
  filter(P_num == 2) %>%
  select(self_pa, partner_pa, self_na, partner_na) %>%
  descr(stats = "common", order = "p") %>%
  knitr::kable(caption = "Descriptives of PANAS for second participants") %>%
  kable_styling()

# csi
csi_all <- descr(baseline$csi_overall, stats = "common", order = "p")
csi_p1  <- descr(baseline[baseline$P_num == 1, "csi_overall"], 
                          stats = "common", order = "p")
csi_p2  <- descr(baseline[baseline$P_num == 2, "csi_overall"], 
                          stats = "common", order = "p")
csi_descr <- cbind(csi_all, csi_p1, csi_p2) %>%
  as.data.frame()
names(csi_descr) <- c("all", "P1", "P2")

csi_descr %>%
  knitr::kable(caption = "Descriptives of Relationship Quality") %>%
  kable_styling()
```

```{r scale-plots}
# BFAS and PANAS
for(ivar in paste("Self", c("Agreeableness", "Conscientiousness", "Extraversion",
                            "Neuroticism", "Openness Domain",
                            "Positive Affect", "Negative Affect"))){
  
  # extract scale name from dict
  var_name <- unique(dict[dict$scale == ivar, "scale_label"])
  
  # make plot
  p <- plot_hist(var = var_name, 
                 var_name = ivar, 
                 data = baseline, bin_width = 0.1) +
    xlim(1,5)

  if(signif(var = var_name,
            data = baseline)) {
    p <- p + labs(subtitle = "Significant participant difference (p < 0.05)")
  }
  print(p)
}

# relationship quality plot
p <- plot_hist(var = "csi_overall", 
               var_name = "Relationship Quality", 
               data = baseline, bin_width = 5)

if(signif(var = "csi_overall",
          data = baseline)) {
  p <- p + labs(subtitle = "Significant participant difference (p < 0.05)")
}
print(p)
```

# III. ESM Data

## 1. Scale Items and Labels

```{r scales-esm, results='asis'}
# tipi and csi
for(iscale in c("Agreeableness", "Conscientiousness",
                "Extraversion", "Neuroticism", "Openness",
                "Relationship Quality")) {
  
  # grab scale items in the dictionary
  current_dict <- dict %>% 
    filter(data == "esm" &
           scale == iscale) %>%
    select(var, var_label, value_label, key)
  
  # grab the first time point for each participant
  df_min <- merge(aggregate(time_idx ~ Participant_ID, data = esm, FUN = min),
                  esm, by = c("Participant_ID", "time_idx"), all.x = TRUE)

  # calculate alpha
  current_alpha <- round(
    psych::alpha(df_min[current_dict$var])$total["raw_alpha"], 
    2)

  # print dictionary
  current_dict %>%
    knitr::kable(caption = paste("Scale items for", iscale,
                                 "- Cronbach's alpha =", current_alpha)) %>%
    kable_styling() %>%
    print()
}

# emotions
dict %>%
  filter(var %in% c("mpa", "mna")) %>%
  select(var, var_label, value_label, scale) %>%
  knitr::kable(caption = "Scale items for momentary affects") %>%
  kable_styling()
```

## 2. Descriptives and plots

```{r descr}
esm %>%
  select(tipi_agree, tipi_consci, tipi_extra, tipi_neuro, tipi_open,
         csi_short, mpa, mna, partner_presence) %>%
  descr(stat = "common", order = "p") %>%
  knitr::kable() %>%
  kable_styling()
```

#TODO: add descriptives for completion rates