---
title: "Dynamic Assortative Mating - Analytic Results"
author: "Linh Nguyen"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: 'hide'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, results = 'hide', message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(digits = 3)

libraries <- c("here",         # directory access
               "rties",        # coupled damped oscillator
               "psych",        # fisher's z
               "Hmisc",        # correlation matrix
               "rio",          # import export
               "tidyr",        # long to wide
               "forestplot",   # forest plot of actual vs perceived
               "ggplot2",      # plots
               "kableExtra",   # scroll box
               "dplyr",        # general wrangling
               "lavaan")       # APIM

lapply(libraries, require, character.only = TRUE)

bl_path   <- paste0(here(), "/data/baseline_cleaned.RDS")
esm_path  <- paste0(here(), "/data/esm_cleaned.RDS")
dict_path <- paste0(here(), "/data/dict.xlsx")

source(paste0(here(), "/script/00_helpers.R"))
```

# Meta Data

This is the analytic report for the Dynamic Assortative Mating paper.

- Online access: This analytic report is published on [RPubs](https://rpubs.com/nguyenllpsych/esmdyad-results).
- Preregistration document: [OSF](https://osf.io/t3gnh).
- Descriptives and exploratory data analysis: [RPubs](https://rpubs.com/nguyenllpsych/esmdyad-eda).

```{r dat}
# load in data
baseline <- rio::import(bl_path)
esm <- rio::import(esm_path)
dict <- rio::import(dict_path)

# example data
head(baseline) %>% 
  knitr::kable(
    caption = "Example first 6 rows of the baseline dataframe"
  ) %>% 
  kableExtra::kable_styling() %>%
  scroll_box(height = "300px")

head(esm %>% filter(Participant_ID == "1002001") %>% arrange(time_idx)) %>%
  knitr::kable(caption = "Example first 6 rows of the esm dataframe") %>%
  kable_styling() %>%
  column_spec(4, width_min = "1in") %>%
  scroll_box(width = "900px")

# retain only analytic variables
baseline <- baseline %>%
  select(Couple_ID, P_num,
         age, sex, gender, sexuality, edu, ses, politics, religion, race_cat,
         panas_self_1:panas_self_20,
         self_neuro:partner_opena)
esm <- esm %>%
  select(Couple_ID, P_num, time_idx, day_idx, mpa, mna, partner_presence,
         tipi_extra:csi_short)
```

# Research Question 1: Evidence of Assortative Mating 

## Hypothesis 1: Baseline Assortative Mating

#### At baseline, romantic partners are similar in their personality traits and general affects such that their scale scores are significantly and positively correlated.

```{r h1}
# run function
h1_results <- h1_function(
  var_list = c(paste0("self_", 
                      c("pa", "na",
                        "agree", "compa", "polit", 
                        "consci", "indus", "order", 
                        "extra", "assert", "enthu", 
                        "neuro", "volat", "withd", 
                        "open", "intel", "opena"))),
  prof_list = data.frame(
    pa      = dict %>% filter(scale == "Self Positive Affect") %>% pull(var),
    na      = dict %>% filter(scale == "Self Negative Affect") %>% pull(var),
    panas   = paste0("panas_self_", 1:20),
    domains = paste0("self_", c("agree", "consci", "extra", "neuro", "open")),
    aspects = paste0("self_", c("compa", "polit", "indus", "order", "assert",
                                "enthu", "volat", "withd", "intel", "opena"))),
  .data = baseline)

# merge profile correlations back to dataframe
baseline <- merge(baseline, h1_results$profile_df, all = TRUE)

# print results
h1_results$bivariate %>% 
  knitr::kable(caption = "Bivariate between-partner correlation") %>% 
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h1_results$profile %>% 
  knitr::kable(
    caption = "Proportion of signficant between-partner profile correlations") %>% 
  kable_styling("striped")
```

## Hypothesis 2: Perceived vs. Actual

#### At baseline, perceived similarity is stronger than actual similarity

```{r h2, fig.height = 10}
# run function
h2_results <- h2_function(
  perception_list = c("pa", "na",
                      "agree", "compa", "polit", 
                      "consci", "indus", "order", 
                      "extra", "assert", "enthu", 
                      "neuro", "volat", "withd", 
                      "open", "intel", "opena"),
  .data = baseline
)

# view results
h2_results$similarity_df %>%
  knitr::kable(
    caption = "Actual and perceived similarities as bivariate correlations"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h2_results$compare_df %>%
  knitr::kable(
    caption = "Comparison between actual and perceived similarities"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

# plot
similarity_df <- h2_results$similarity_df
similarity_df$personality <- rep(
  c(
    "Positive Affect", "Negative Affect",
    "Agreeableness", "Compassion", "Politeness",
    "Conscientiousness", "Industriousness", "Orderliness",
    "Extraversion", "Assertiveness", "Enthusiasm",
    "Neuroticism", "Volatility", "Withdrawal",
    "Openness", "Intellect", "Openness Aspect"),
  each = 3)
plot_forest(
  perception_list = c("pa", "na",
                      "agree", "compa", "polit", 
                      "consci", "indus", "order", 
                      "extra", "assert", "enthu", 
                      "neuro", "volat", "withd", 
                      "open", "intel", "opena"),
  similarity_df = similarity_df)
```

## Hypothesis 3: Dynamic Assortative Mating

#### Dynamically, romantic partners are similar in their momentary variability on personality and affective states.

```{r h3, warning = F}
# run function
h3_results <- h3_function(
  var_list = c("mpa", "mna", 
               paste0("tipi_", c("agree", "consci", "extra", "neuro", "open"))),
  .data = esm,
  .dict = dict %>% filter(data == "esm")
)

# merge within-person indices back to baseline dataframe
baseline <- merge(baseline,
                  h3_results$within_df)

# show results
h3_results$cor_tab %>% 
  knitr::kable(
    caption = "Dynamic Between-Partner Correlations") %>% 
  kable_styling("striped") %>%
  scroll_box(width = "900px")
```

# Research Question 2: Benefits of Assortative Mating

## Hypothesis 4: Baseline Benefits

#### At baseline, partner similarity in self-reported personality traits and general affect is associated with enhanced relationship quality

```{r h4}
# run function
h4_results <- h4_function(
  var_list = c(paste0("self_", 
                      c("pa", "na",
                        "agree", "compa", "polit", 
                        "consci", "indus", "order", 
                        "extra", "assert", "enthu", 
                        "neuro", "volat", "withd", 
                        "open", "intel", "opena"))),
  prof_list = c("pa", "na", "panas", "domains", "aspects"),
  quality_var = "csi_overall",
  .data = baseline
)

# results
h4_results$interaction_tab %>%
  knitr::kable(
    caption = "Multiple Regression Models with Interaction Effects"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h4_results$difference_tab %>%
  knitr::kable(
    caption = "Simple Regression Model with Difference Score"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h4_results$profile_tab %>%
  knitr::kable(
    caption = "Simple Regression Model with Profile Similarity"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")
```

## Hypothesis 5: APIM Actual Similarities

```{r h5, warning = F}
# run function
h5_results <- h5_function(
  var_list = c("pa", "na",
               "agree", "compa", "polit", 
               "consci", "indus", "order", 
               "extra", "assert", "enthu", 
               "neuro", "volat", "withd", 
               "open", "intel", "opena"),
  quality_var = "csi_overall",
  .data = baseline
)

# view results
h5_results$actual %>%
  knitr::kable(
    caption = "Standardized solutions for APIM models with actual similarity"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h5_results$perceived_p1 %>%
  knitr::kable(
    caption = "Standardized solutions for APIM models with P1-perceived similarity"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")
h5_results$perceived_p2 %>%
  knitr::kable(
    caption = "Standardized solutions for APIM models with P2-perceived similarity"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")
```

## Hypothesis 6: APIM Perceived Similarities

```{r h6, warning = F}
# output included in h5_function
h5_results$perceived_p1 %>%
  knitr::kable(
    caption = "Standardized solutions for APIM models with P1-perceived similarity"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")
h5_results$perceived_p2 %>%
  knitr::kable(
    caption = "Standardized solutions for APIM models with P2-perceived similarity"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")
```

## Hypothesis 7: Dynamic Benefits

#### Dynamically, partner similarity in momentary variability on personality and affective states is associated with enhanced relationship quality.

```{r h7}
# run function 
# same as h4 but without the profile argument
h7_results <- h4_function(
  var_list = apply(expand.grid(
    c("mpa", "mna", 
      "tipi_agree", "tipi_consci", "tipi_extra", "tipi_neuro", "tipi_open"),
    c("_sd", "_rvi", "_mssd")), 1, paste, collapse = ""),
  quality_var = "csi_overall",
  .data = baseline
)

### z-correlation as predictors ###
fluctuation_tab <- data.frame()

# sort data
fluctuation_dat <- baseline %>% arrange(Couple_ID)

#   extract vector of each partner score on relationship quality variables
p1_qual <- fluctuation_dat[fluctuation_dat$P_num == 1, "csi_overall", drop = T]
p2_qual <- fluctuation_dat[fluctuation_dat$P_num == 2, "csi_overall", drop = T]

for(ivar in paste0(
  c("mpa", "mna", 
    "tipi_agree", "tipi_consci", "tipi_extra", "tipi_neuro", "tipi_open"),
  "_z")) {
  
  # extract vector of fisher's z-transformed within-couple correlation
  #   same for both participants, use P_num = 1 to avoid duplication
  z_var <- fluctuation_dat[fluctuation_dat$P_num == 1, ivar, drop = T]
  
  # fit simple regression models
  mod_1 <- summary(lm(p1_qual ~ z_var))$coefficients
  mod_2 <- summary(lm(p2_qual ~ z_var))$coefficients
  
  # extract coefficients
  z_1   <- round(mod_1["z_var", 
                       c("Estimate", "t value", "Pr(>|t|)")], 3)
  z_2   <- round(mod_2["z_var", 
                       c("Estimate", "t value", "Pr(>|t|)")], 3)
  
  # store values in difference_tab
  fluctuation_tab <- rbind(
    fluctuation_tab,
    
    # first partner
    c(ivar, "P1", z_1),
    
    # second partner
    c(ivar, "P2", z_2)
    )
} # END for ivar LOOP

#   rename results table
names(fluctuation_tab) <- c("personality", "participant",
                            "z_est", "z_tval", "z_pval")

# view results
h7_results$interaction_tab %>%
  knitr::kable(
    caption = "Multiple Regression Models with Interaction Effects"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h7_results$difference_tab %>%
  knitr::kable(
    caption = "Simple Regression Model with Difference Score"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

fluctuation_tab %>%
  knitr::kable(
    caption = "Simple Regression Model with Fluctuations"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")
```

# Exploratory Model: Coupled Damped Oscillator

```{r rties}
# make sure all time points are present
template <- expand.grid(Couple_ID = unique(esm$Couple_ID),
                        P_num = 1:2,
                        time_idx = 0:34)

# left join the template with esm so all missing rows are present
rties_dat <- left_join(template, esm, by = c("Couple_ID", "P_num", "time_idx"))

# drop duplicates if any
rties_dat <- rties_dat[!duplicated(rties_dat), ]

# prep data according to rties instructions
rties_dat$couple <- rties_dat$Couple_ID - 1000
rties_dat$person <- ifelse(rties_dat$P_num == 1, 
                           rties_dat$couple, rties_dat$couple + 500)

# remove duplicated data at same time_idx
rties_dat <- rties_dat[!duplicated(rties_dat[c("Couple_ID", "P_num", "time_idx")])| 
                        duplicated(rties_dat[c("Couple_ID", "P_num", "time_idx")], 
                                   fromLast = TRUE), ]

# fix time_idx to start at 1
rties_dat$time_idx <- rties_dat$time_idx + 1

# use rties::dataPrep()
prepped_dat <- dataPrep(basedata = rties_dat, 
                        dyadId = "couple", personId = "person",
                        obs_name = "tipi_neuro", dist_name = "P_num",
                        time_name = "time_idx")


## test taus, embeds, and delta
## not successful
#taus <- c(4,5)
#embeds <- c(3,4,5)
#delta <- 1
#
#derivs <- estDerivs(prepData=prepped_dat, taus=taus, embeds=embeds, delta=delta, 
#                    idConvention=500)

# random selection of couples with at least 28 shared datapoints
set.seed(202403)
id_list <- baseline %>% filter(tipi_open_n > 28) %>%
  pull(Couple_ID) %>%
  unique()
id_random <- sample(id_list, size = 9)

# update time to hour unit
old_time <- 1:35
new_time <- c(seq(from = 0, to = 12, by = 3),
              seq(from = 0, to = 12, by = 3)+24*1,
              seq(from = 0, to = 12, by = 3)+24*2,
              seq(from = 0, to = 12, by = 3)+24*3,
              seq(from = 0, to = 12, by = 3)+24*4,
              seq(from = 0, to = 12, by = 3)+24*5,
              seq(from = 0, to = 12, by = 3)+24*6)
rties_dat$new_time <- NA
for (i in seq_along(old_time)) {
  # Update the corresponding value in time with the value from new_time
  rties_dat$new_time[rties_dat$time_idx == old_time[i]] <- new_time[i]
}
rties_dat$time_idx <- rties_dat$new_time

# plot random couples
plot_timeseries(basedata=rties_dat %>% filter(Couple_ID %in% id_random), 
                dyad="couple", 
                obs_name="mna", 
                dist_name="P_num", 
                time_name="time_idx", 
                dist0name="P2", dist1name= "P1", 
                plot_obs_name="Negative Affect", printPlots = F)
plot_timeseries(basedata=rties_dat %>% filter(Couple_ID %in% id_random), 
                dyad="couple", 
                obs_name="mpa", 
                dist_name="P_num", 
                time_name="time_idx", 
                dist0name="P2", dist1name= "P1", 
                plot_obs_name="Positive Affect", printPlots = F)
plot_timeseries(basedata=rties_dat %>% filter(Couple_ID %in% id_random), 
                dyad="couple", 
                obs_name="tipi_agree", 
                dist_name="P_num", 
                time_name="time_idx", 
                dist0name="P2", dist1name= "P1", 
                plot_obs_name="Agreeableness", printPlots = F)
plot_timeseries(basedata=rties_dat %>% filter(Couple_ID %in% id_random), 
                dyad="couple", 
                obs_name="tipi_consci", 
                dist_name="P_num", 
                time_name="time_idx", 
                dist0name="P2", dist1name= "P1", 
                plot_obs_name="Conscientiousness", printPlots = F)
plot_timeseries(basedata=rties_dat %>% filter(Couple_ID %in% id_random), 
                dyad="couple", 
                obs_name="tipi_extra", 
                dist_name="P_num", 
                time_name="time_idx", 
                dist0name="P2", dist1name= "P1", 
                plot_obs_name="Extraversion", printPlots = F)
plot_timeseries(basedata=rties_dat %>% filter(Couple_ID %in% id_random), 
                dyad="couple", 
                obs_name="tipi_neuro", 
                dist_name="P_num", 
                time_name="time_idx", 
                dist0name="P2", dist1name= "P1", 
                plot_obs_name="Neuroticism", printPlots = F)
plot_timeseries(basedata=rties_dat %>% filter(Couple_ID %in% id_random), 
                dyad="couple", 
                obs_name="tipi_open", 
                dist_name="P_num", 
                time_name="time_idx", 
                dist0name="P2", dist1name= "P1", 
                plot_obs_name="Openness", printPlots = F)
plot_timeseries(basedata=rties_dat %>% filter(Couple_ID %in% id_random), 
                dyad="couple", 
                obs_name="csi_short", 
                dist_name="P_num", 
                time_name="time_idx", 
                dist0name="P2", dist1name= "P1", 
                plot_obs_name="Relationship Satisfaction", printPlots = F)
```

