---
title: "Dynamic Assortative Mating - Analytic Results"
author: "Linh Nguyen"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: 'hide'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, results = 'hide', message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
options(digits = 3)

libraries <- c("here",         # directory access
               "psych",        # fisher's z
               "Hmisc",        # correlation matrix
               "rio",          # import export
               "tidyr",        # long to wide
               "forestplot",   # forest plot of actual vs perceived
               "ggplot2",      # plots
               "kableExtra",   # scroll box
               "dplyr",        # general wrangling
               "lavaan")       # APIM

lapply(libraries, require, character.only = TRUE)

bl_path   <- paste0(here(), "/data/2024-01-15/baseline_cleaned_2024-01-15.RDS")
esm_path  <- paste0(here(), "/data/2024-01-15/esm_cleaned_2024-01-15.RDS")
dict_path <- paste0(here(), "/data/dict.xlsx")

source(paste0(here(), "/script/00_helpers.R"))
```

# Meta Data

This is the analytic report for the Dynamic Assortative Mating paper.

- Online access: This analytic report is published on [RPubs](https://rpubs.com/nguyenllpsych/esmdyad-results).
- Preregistration document: [OSF](https://osf.io/t3gnh).
- Descriptives and exploratory data analysis: [RPubs](https://rpubs.com/nguyenllpsych/esmdyad-eda).

```{r dat}
# load in data
baseline <- rio::import(bl_path)
esm <- rio::import(esm_path)
dict <- rio::import(dict_path)

# example data
head(baseline) %>% 
  knitr::kable(
    caption = "Example first 6 rows of the baseline dataframe"
  ) %>% 
  kableExtra::kable_styling() %>%
  scroll_box(height = "300px")

head(esm %>% filter(Participant_ID == "1002001") %>% arrange(time_idx)) %>%
  knitr::kable(caption = "Example first 6 rows of the esm dataframe") %>%
  kable_styling() %>%
  column_spec(4, width_min = "1in") %>%
  scroll_box(width = "900px")

# retain only analytic variables
baseline <- baseline %>%
  select(Couple_ID, P_num,
         age, sex, gender, sexuality, edu, ses, politics, religion, race_cat,
         panas_self_1:panas_self_20,
         self_neuro:partner_opena)
esm <- esm %>%
  select(Couple_ID, P_num, time_idx, day_idx, mpa, mna, partner_presence,
         tipi_extra:csi_short)
```

# Hypothesis 1: Baseline Assortative Mating

#### At baseline, romantic partners are similar in their personality traits and general affects such that their scale scores are significantly and positively correlated.

```{r h1}
# run function
h1_results <- h1_function(
  var_list = c(paste0("self_", 
                      c("pa", "na",
                        "agree", "compa", "polit", 
                        "consci", "indus", "order", 
                        "extra", "assert", "enthu", 
                        "neuro", "volat", "withd", 
                        "open", "intel", "opena"))),
  prof_list = data.frame(
    pa      = dict %>% filter(scale == "Self Positive Affect") %>% pull(var),
    na      = dict %>% filter(scale == "Self Negative Affect") %>% pull(var),
    panas   = paste0("panas_self_", 1:20),
    domains = paste0("self_", c("agree", "consci", "extra", "neuro", "open")),
    aspects = paste0("self_", c("compa", "polit", "indus", "order", "assert",
                                "enthu", "volat", "withd", "intel", "opena"))),
  .data = baseline)

# merge profile correlations back to dataframe
baseline <- merge(baseline, h1_results$profile_df, all = TRUE)

# print results
h1_results$bivariate %>% 
  knitr::kable(caption = "Bivariate between-partner correlation") %>% 
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h1_results$profile %>% 
  knitr::kable(
    caption = "Proportion of signficant between-partner profile correlations") %>% 
  kable_styling("striped")
```

# Hypothesis 2: Dynamic Assortative Mating

#### Dynamically, romantic partners are similar in their momentary variability on personality and affective states.

```{r h2}
# run function
h2_results <- h2_function(
  var_list = c("mpa", "mna", 
               paste0("tipi_", c("agree", "consci", "extra", "neuro", "open"))),
  .data = esm,
  .dict = dict
)

# merge within-person indices back to baseline dataframe
baseline <- merge(baseline,
                  h2_results$within_df)

# show results
h2_results$cor_tab %>% 
  knitr::kable(
    caption = "Dynamic Between-Partner Correlations") %>% 
  kable_styling("striped") %>%
  scroll_box(width = "900px")
```

# Hypothesis 3: Baseline Benefits

#### At baseline, partner similarity in self-reported personality traits and general affect is associated with enhanced relationship quality

```{r h3}
# run function
h3_results <- h3_function(
  var_list = c(paste0("self_", 
                      c("pa", "na",
                        "agree", "compa", "polit", 
                        "consci", "indus", "order", 
                        "extra", "assert", "enthu", 
                        "neuro", "volat", "withd", 
                        "open", "intel", "opena"))),
  prof_list = c("pa", "na", "panas", "domains", "aspects"),
  quality_var = "csi_overall",
  .data = baseline
)

# results
h3_results$interaction_tab %>%
  knitr::kable(
    caption = "Multiple Regression Models with Interaction Effects"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h3_results$difference_tab %>%
  knitr::kable(
    caption = "Simple Regression Model with Difference Score"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h3_results$profile_tab %>%
  knitr::kable(
    caption = "Simple Regression Model with Profile Similarity"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

```

# Hypothesis 4: Dynamic Benefits

#### Dynamically, partner similarity in momentary variability on personality and affective states is associated with enhanced relationship quality.

```{r h4}
# run function 
# same as h3 but without the profile argument
h4_results <- h3_function(
  var_list = baseline %>% select(mpa_sd:tipi_open_mssd) %>% names(),
  quality_var = "csi_overall",
  .data = baseline
)

# view results
h4_results$interaction_tab %>%
  knitr::kable(
    caption = "Multiple Regression Models with Interaction Effects"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h4_results$difference_tab %>%
  knitr::kable(
    caption = "Simple Regression Model with Difference Score"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")
```

# Hypothesis 5: APIM

```{r h5, warning = F}
# run function
h5_results <- h5_function(
  var_list = c("pa", "na",
               "agree", "compa", "polit", 
               "consci", "indus", "order", 
               "extra", "assert", "enthu", 
               "neuro", "volat", "withd", 
               "open", "intel", "opena"),
  quality_var = "csi_overall",
  .data = baseline
)

# view results
h5_results$actual %>%
  knitr::kable(
    caption = "Standardized solutions for APIM models with actual similarity"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h5_results$perceived_p1 %>%
  knitr::kable(
    caption = "Standardized solutions for APIM models with P1-perceived similarity"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")
h5_results$perceived_p2 %>%
  knitr::kable(
    caption = "Standardized solutions for APIM models with P2-perceived similarity"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")
```

# Hypothesis 6: Perceived vs. Actual

#### At baseline, perceived similarity is stronger than actual similarity, and it is in turn more strongly associated with relationship quality

```{r h6, fig.height = 10}
# run function
h6_results <- h6_function(
  perception_list = c("pa", "na",
                      "agree", "compa", "polit", 
                      "consci", "indus", "order", 
                      "extra", "assert", "enthu", 
                      "neuro", "volat", "withd", 
                      "open", "intel", "opena"),
  .data = baseline
)

# view results
h6_results$similarity_df %>%
  knitr::kable(
    caption = "Actual and perceived similarities as bivariate correlations"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

h6_results$compare_df %>%
  knitr::kable(
    caption = "Comparison between actual and perceived similarities"
  ) %>%
  kable_styling("striped") %>%
  scroll_box(height = "300px")

# plot
similarity_df <- h6_results$similarity_df
similarity_df$personality <- rep(
  c(
    "Positive Affect", "Negative Affect",
    "Agreeableness", "Compassion", "Politeness",
    "Conscientiousness", "Industriousness", "Orderliness",
    "Extraversion", "Assertiveness", "Enthusiasm",
    "Neuroticism", "Volatility", "Withdrawal",
    "Openness", "Intellect", "Openness Aspect"),
  each = 3)
plot_forest(
  perception_list = c("pa", "na",
                      "agree", "compa", "polit", 
                      "consci", "indus", "order", 
                      "extra", "assert", "enthu", 
                      "neuro", "volat", "withd", 
                      "open", "intel", "opena"),
  similarity_df = similarity_df)
```

